Addition and removal energies of circular quantum dots

### 2017-06-13 Manuscript/paper.tex
changed paper.tex

\begin{abstract}
  We present and compare several many-body methods as applied to two-dimensional quantum dots with circular symmetry.  We calculate the approximate ground state energy using a harmonic oscillator basis optimized by Hartree--Fock (HF) theory and further improve the ground state energy using one of two post-HF methods: in-medium similarity renormalization group (IM-SRG) and coupled cluster singles-and-doubles (CCSD).  With the application of quasidegenerate perturbation theory (QDPT) or the equation-of-motion (EOM) method, we obtain addition and removal energies as well.  Our results are benchmarked against full configuration interaction (FCI) and diffusion Monte Carlo (DMC) where available.  We examine the rate of convergence and perform extrapolations to the infinite basis limit using a power-law model.
\end{abstract}

@@

The behavior of strongly confined electrons is of fundamental concern to many-body theory.  The prototypical system is that of quantum dots, also known as ``artificial atoms'', in which electrons are confined within artificial constructed semiconducting heterostructures.  Such nanoscale systems are highly relevant as they lie at a critical point where both theoretical models and experimental observations can readily probe quantum phenomena such as tunneling, entanglement, and magnetization \cite{reimann2002,engel1993}.  Moreover, unlike many physical systems, quantum dots enjoy the benefit of being highly tunable through changes in the external field or the structure of the confining material.  This allows experiments to quantify the impact of quantum effects at different levels of correlation.

The ground states of quantum dots exhibit shell structures and magic numbers similar to those of atoms and nuclei \cite{tarucha1996}.  Thus, they provide a rare opportunity to study electronic systems without the influence of the atomic nuclei.  They also provide a testbed for the development of many-body methods for other systems with similar geometry, such as nuclei or neutrons drops.

Beyond their theoretical relevance, there are numerous potential applications quantum dots.  The electrical and optical properties of quantum dots are particularly useful for solar cells \cite{jenks:013111,doi:10.1021/cr900289f} and laser technology \cite{strauf2010,5075760}.  They are have potential medical applications in diagnosis and treatment \cite{Ben-Ari02042003}.  And of course, quantum dots are also promising candidates for the physical realization of quantum computing\cite{PhysRevA.57.120}.

In general, quantum dot systems are not analytically solvable, with the exception of 2-electron systems or systems with specific strengths of the external field \cite{PhysRevA.48.3561}.  Hence, in practical applications one can only hope for a numerical solution through some \textit{many-body method} or a combination thereof.

One of the most accurate techniques for solving many-body systems is that of exact diagonalization, also known as the full configuation interaction method.  Unlike most methods, it is assured to converge to the exact answer as the size of the finite basis is increased to infinity (the so-called \textit{infinite-basis limit}).  Despite this significant advantage, it is often infeasible to perform exact diagonalization due to its factorially increasing cost with respect to the basis size and the number of particles.  This led to the development of more cost-effective methods that trade varying amounts of accuracy for varying amounts of speed.

In this paper, we analyze the application of several methods to the quantum dot systems, including the Hartree--Fock (HF) method, M\o ller-Plesset perturbation theory (MP), the in-medium similarity renormalization group (IM-SRG) method, coupled cluster (CC) theory \cite{PhysRevB.67.045320,heidari:114708,PhysRevB.84.115302}, quasidegenerate perturbation theory (QDPT), and equation-of-motion (EOM) methods.  We compare them against variational and diffusion Monte Carlo (VMC and DMC) \cite{PhysRevB.68.035304,PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780} and full configuration interaction (FCI) \cite{olsen2013thesis,JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102} results from existing literature.

@@

The article is organized as follows: Section \ref{sec:formalism} introduces first (\ref{subsec:modelHamiltonian}) the Hamiltonian we use to model circular quantum dots, and gives afterwards an overview of the \textit{ab initio} many-body methods used in this paper: Hartree--Fock, IM-SRG, CCSD, QDPT, EOM.  Our results are presented in Section \ref{sec:results}.  We analyze and compare the differences between the various methods and also between different quantum dot systems.  We examine the usefulness extrapolation techniques in improving the precision of the results.  Section \ref{sec:conclusions} concludes our work and gives perspectives for future work.

@@

We shall model the circular quantum dot system as a collection of $N$ nonrelativistic electrons of mass $m$ in 2-dimensional space, trapped by an external harmonic-oscillator potential of the form $m \omega^2 r^2 / 2$, where $\omega$ is the angular frequency of the oscillator trap and $r$ is the distance from the center of the trap.  The electrons interact with each other through the standard Coulomb interaction $e^2 / (4 \pi \epsilon R)$, where $e$ is the charge of each electron, $\epsilon$ is the permittivity of the medium, and $R$ is the distance between the two interacting electrons.

Even though the model system contains 5 parameters $(N, m, \omega, e, \epsilon)$, many of them are redundant.  With an appropriate choice of units, the number of parameters in the system can be reduced to just two.  For rest of this paper, we shall use atomic units\footnote{Here, ``atomic unit'' is used in an \emph{abstract} sense: we do not assume (nor care) whether $m$, $e$, $\epsilon$ correspond to true values in vacuum, or some effective value in medium.} where $\hbar$, $m$, $e$ and $4 \pi \epsilon$ are all set to one.  Hartree $E_{\mathrm{h}}$ becomes the unit of energy and the Bohr radius $a$ becomes the unit of length:
\begin{align*}
  E_{\mathrm{h}} &\equiv m \left(\frac{e^2}{4 \pi \epsilon \hbar}\right)^2 &
  a &\equiv \frac{4 \pi \epsilon \hbar^2}{m e^2}
\end{align*}
This leaves us with $(N, \omega)$ as the only parameters necessary to specify the quantum dot system, with $\omega$ in units of $E_{\mathrm{h}} / \hbar$.

The behavior of each electron in isolation is described by the nonrelativistic kinetic energy $t(\bm{r})$ and the potential energy due to the trap $u(\bm{r})$, which combine to form the standard harmonic oscillator Hamiltonian $h(\bm{p}, \bm{r})$:
\begin{align*}
  t(\bm{p}) &\equiv \frac{1}{2} \bm{p}^2 &
  u(\bm{r}) &\equiv \frac{1}{2} \omega^2 \bm{r}^2 &
  h(\bm{p}, \bm{r}) &\equiv t(\bm{p}) + t(\bm{r})
\end{align*}
Here, $\bm{p}$ is the linear momentum of the electron and $\bm{r}$ is its position relative to the center of the trap.  The pairwise interaction between two electrons at positions $\bm{r}$ and $\bm{r}'$ is simply the Coulomb interaction:
\begin{align*}
  v(\bm{r}, \bm{r}') \equiv \frac{1}{|\bm{r} - \bm{r}'|}
\end{align*}
Then, the full many-body problem is described by the Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{[1]} + \hat H^{[2]}
\end{align*}
where $\hat{H}^{[1]}$ and $\hat{H}^{[2]}$ are the one- and two-body parts of the Hamiltonian,
\begin{align} \label{eq:onetwobodyhamiltonian}
\hat{H}^{[1]} &\equiv \sum_{j = 1}^N h(\hat{\bm p}_j, \hat{\bm r}_j) &
\hat{H}^{[2]} &\equiv \sum_{j_1 = 1}^N \sum_{j_2 = 1}^{j_1 - 1} v(\hat{\bm r}_{j_1}, \hat{\bm r}_{j_2})
\end{align}
The operator $\hat{\bm r}_j$ is the position operator of the $j$-th particle,\footnote{The ordering of the particle labels $j$ is unimportant as they exist only for the purpose of bookkeeping.} and $\hat{\bm p}_j \equiv -\mathrm{i} \hat{\bm{\nabla}}_{\bm{r}_j}$ is its momentum operator.

@@

The noninteracting part of the many-body Hamiltonian $\hat{H}^{[1]}$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  In Cartesian coordinates, the two-dimensional harmonic oscillator is trivially reducible to the well-known one-dimensional problem.  However, to exploit the circular symmetry, we prefer to use Fock-Darwin states, which are written in polar coordinates $\bm{r} = (\rho, \varphi)$.  Such states have the computationally useful property of conserving orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv -\mathrm{i} \partial / (\partial \varphi)$.

The Fock-Darwin wave functions can be decomposed into radial and angular components,\cite[Appx.\ A]{lohne2010coupled}
\begin{align}
  F_{n, m_\ell}(\rho, \varphi) &\equiv \sqrt\omega R_{n, |m_\ell|}(\sqrt \omega \rho) A_{m_\ell}(\varphi) \label{eq:fockdarwin} \\
  R_{n, m}(\varrho) &\equiv \sqrt{\frac{2 \cdot n!}{(n + m)!}} \mathrm{e}^{-\varrho^2 / 2} \varrho^m L_n^{(m)}(\varrho^2) \notag \\
  A_{m_\ell}(\varphi) &\equiv \frac{1}{\sqrt{2 \pi}} \mathrm{e}^{\mathrm{i} m_\ell \varphi} \notag
\end{align}
where $L_n^{(\alpha)}$ denotes the (generalized) Laguerre polynomial \cite[\S 18.3]{NIST:DLMF} of degree $n$ and parameter $\alpha$,
\begin{align*}
  L_n^{(\alpha)}(u) = \frac{1}{n!} u^{-\alpha} \mathrm{e}^u \frac{\mathrm{d}^n}{\mathrm{d} u^n} (\mathrm{e}^{-u} u^{\alpha + n})
\end{align*}
The states are labeled by two quantum numbers: the principal quantum number $n \in \{0, 1, 2, \ldots\}$ and orbital angular momentum projection $m_\ell \in \mathbb \{\ldots, 2, -1, 0, 1, 2, \ldots\}$.  For a wave function, $n$ indicates the degree of the Laguerre polynomial, whereas $m_\ell$ is simply the eigenvalue of $\hat{L}_{\mathrm{z}}$.

Since electrons are spin-$1/2$ fermions, they can occupy either of the two possible spin states $\chi_{-1/2}$ or $\chi_{+1/2}$.  Thus, every single-particle basis state $| (n, m_\ell, m_{\mathrm{s}}) \rangle$ contains both a spatial component \eqref{eq:fockdarwin} and a spin component,
\begin{align} \label{eq:singleparticlestate}
  \langle (\rho, \varphi), m_{\mathrm{s}}' | (n, m_\ell, m_{\mathrm{s}}) \rangle \equiv F_{n, m_\ell}(\rho, \varphi) \delta_{m_{\mathrm{s}}^{}, m_{\mathrm{s}}'}
\end{align}
Here we have introduced spin projection $m_{\mathrm{s}} \in \bigl\{-1/2, +1/2\bigr\}$ as the third quantum number and $\delta_{p, q}$ to denote the Kronecker delta..

The energy of the single-particle state $| (n, m_\ell, m_{\mathrm{s}}) \rangle$ is given by
\begin{align} \label{eq:energysingleparticlestate}
  \varepsilon_{n, m_\ell, m_{\mathrm{s}}} \equiv (2 n + |m_\ell| + 1) \omega
\end{align}
The energies are degenerate with respect to the spin projection $m_{\mathrm{s}}$ as our Hamiltonian $\hat{h}$ does not distinguish between them.  Additionally, they are degenerate with respect to the nonnegative integer
\begin{align} \label{eq:shell_index}
  k \equiv 2 n + |m_\ell|
\end{align}
which we conveniently call the \textit{shell index} as it labels each shell starting from zero.  The shells are equidistant with an energy spacing equal to the frequency $\omega$.

Returning to the many-body Hamiltonian $\hat{H}^{[1]}$, we can use the Slater determinant construction to build an $N$-particle eigenstate $|p_1, \ldots, p_N\rangle$ of the many-body Hamiltonian $\hat{H}^{[1]}$ from an arbitrary choice of $N$ single-particle eigenstates $|p_1\rangle, \ldots, |p_N\rangle$ from \eqref{eq:singleparticlestate}, where $p$ is an abbreviation for the triplet $(n, m_\ell, m_{\mathrm{s}})$.  Note that a Slater determinant $|p_1, \ldots, p_N\rangle$ is not merely a product of the single-particle states, but an antisymmetrized sum over all possible products,
\begin{align*}
  \langle \bm{r}_1, \ldots, \bm{r}_N | p_1, \ldots, p_N \rangle \equiv
  \frac{1}{\sqrt{N}} \left|
  \begin{matrix}
    \langle \bm{r}_1 | p_1 \rangle & \cdots & \langle \bm{r}_1 | p_N \rangle \\
    \vdots & \ddots & \vdots \\
    \langle \bm{r}_N | p_1 \rangle & \cdots & \langle \bm{r}_N | p_N \rangle \\
  \end{matrix}
  \right|
\end{align*}
The Slater determinant $|p_1, \ldots, p_N\rangle$ is said to \textit{occupy} the states $|p_1\rangle, \ldots, |p_N\rangle$.  By construction, Slater determinants satisfy the Pauli exclusion principle, ensuring that the exchange of any two particles is equivalent to a reversal in the sign of the many-body wave function.

@@

Currently, the Hamiltonian $\hat{H}$ is defined for a fixed number of particles $N$.  Through the formalism of second quantization, the many-body machinery can be made agnostic to number of particles, in analogy to quantum field theory.  Instead of considering the Hilbert space for exactly $N$ particles, we combine every space with $N = 0, 1, 2, \ldots$ particles into a combined Hilbert space called the Fock space.  To establish relationships between states with differing number of particles, we introduce the \textit{field operators} $\hat{a}_p^\dagger$ and $\hat{a}_p$ to increment and decrement the number of particles in a state, respectively.  The linear operator $\hat{a}_p^\dagger$ is known as \textit{creation operator} for the single-particle state $p$.  Its action on an arbitrary Slater determinant is defined as:
\begin{align*}
  \hat a_p^\dagger |p'_1, \ldots, p'_N\rangle =
  |p, p'_1, \ldots, p'_N\rangle
\end{align*}
Informally, it is said to \textit{create} a particle in the state $p$.  If $p$ coincides with any of the existing $p'_1, \ldots, p'_N$, then antisymmetrization would collapse the state to the number $0$, an invalid state.  The Hermitian conjugate of $\hat{a}_p^\dagger$ is $\hat{a}_p$, which known as the \textit{annihilation operator} for the state $p$.  By analogy, its action is to remove the state $p$ from a Slater determinant, or collapse it to the number $0$ if the state $p$ was not occupied in the Slater determinant.

From here, one can show that Slater determinants themselves can be built directly from the the unique \textit{vacuum state} $|\rangle$ with no particles ($N = 0$):
\begin{align*}
  |p_1, \ldots, p_N\rangle =
  \hat a_{p_1}^\dagger \cdots \hat a_{p_N}^\dagger |\rangle
\end{align*}
To properly describe fermions, the field operators $\hat a_p$ and $\hat a_p^\dagger$ must satisfy the anticommutation relations that encode the Pauli principle,
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0 &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p, q}
\end{align*}

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from \eqref{eq:onetwobodyhamiltonian} can be rewritten as\footnote{The peculiar ordering of $\hat a_s \hat a_r$ here is not a typographical error.}
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{p, q} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{} &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{p, q, r, s} \langle p, q | \hat{H}^{[2]} | r, s \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}
\end{align}
where the numeric coefficients $\langle p | \hat{H}^{[1]} | q \rangle$ and $\langle p, q | \hat{H}^{[2]} | r, s \rangle$ are known as \textit{matrix elements} and fully characterize the two operators in the basis.  In this paper, we use the letters $p, q, r, s, t, \ldots$ to label arbitrary single-particle \textit{states} in the basis.  For the harmonic oscillator basis, labels such as $p$ can be considered abbreviations for triplets of the form $(n, m_\ell, m_{\mathrm{s}})$.  A summation such as $\sum_p$ is then understood as a summation over every single-particle basis state $p$.  This is not to be confused with summations such as $\sum_j$ from \eqref{eq:onetwobodyhamiltonian}, which contains a label $j$ denoting an arbitrary \textit{particle}.  We thus see another benefit with second quantization: the Hamiltonian representation makes no attempt to distinguish particles at all, as expected for particles with quantum statistics.

>>>>

\begin{abstract}
  We present and compare several many-body methods as applied to
  two-dimensional quantum dots with circular symmetry.  We calculate
  the approximate ground state energy using a harmonic oscillator
  basis optimized by Hartree--Fock (HF) theory and further improve the
  ground state energy using two post-HF methods: in-medium
  similarity renormalization group (IM-SRG) and coupled cluster
  singles-and-doubles (CCSD).  With the application of quasidegenerate
  perturbation theory (QDPT) or the equation-of-motion (EOM) method,
  we obtain addition and removal energies as well.  Our results are
  benchmarked against full configuration interaction (FCI) and
  diffusion Monte Carlo (DMC) where available.  We examine the rate of
  convergence and perform extrapolations to the infinite basis limit
  using a power-law model.
\end{abstract}

@@

The behavior of strongly confined electrons is of fundamental concern
to many-body theory.  The prototypical system is that of quantum dots,
also known as ``artificial atoms'', in which electrons are confined
within artificially constructed semiconducting heterostructures.  Such
nanoscale systems are highly relevant as they lie at a critical point
where both theoretical models and experimental observations can
readily probe quantum phenomena such as tunneling, entanglement, and
magnetization \cite{reimann2002,engel1993}.  Moreover, unlike many
physical systems, quantum dots enjoy the benefit of being highly
tunable through changes in the external field or the structure of the
confining material.  This allows experiments to quantify the impact of
quantum effects at different levels of correlation.

The ground states of quantum dots exhibit shell structures and magic
numbers similar to those of atoms and nuclei \cite{tarucha1996}.
Thus, they provide a rare opportunity to study electronic systems
without the influence of the atomic nuclei.  They also provide a
testbed for the development of many-body methods for other systems
with similar geometry, such as nuclei or neutrons drops \cite{bogner2010}.

Beyond their theoretical relevance, there are numerous potential
applications of quantum dots.  The electrical and optical properties of
quantum dots are particularly useful for solar
cells \cite{jenks:013111,doi:10.1021/cr900289f} and laser
technology \cite{strauf2010,5075760}.  They have potential medical
applications in diagnosis and treatment \cite{Ben-Ari02042003}.  And
of course, quantum dots are also promising candidates for the physical
realization of quantum computing \cite{PhysRevA.57.120}.

In general, quantum dot systems are not analytically solvable, with
the exception of two-electron systems or systems with specific strengths
of the external field \cite{PhysRevA.48.3561}.  Hence, in practical
applications one can only hope for a numerical solution through
some \textit{many-body method} or a combination thereof.

One of the most accurate techniques for solving many-body systems is
that of exact diagonalization, also known as the full configuration
interaction method.  Unlike most methods, it is assured to converge to
the exact answer as the size of the finite basis is increased to
infinity (the so-called \textit{infinite-basis limit}).  Despite this
significant advantage, it is often not possible  to perform exact
diagonalization due to its factorially increasing cost with respect to
the basis size and the number of particles.  This led to the
development of more cost-effective methods that trade varying amounts
of accuracy for varying amounts of speed.

In this paper, we analyze the application of several methods to the
quantum dot systems, including the Hartree--Fock (HF) method, M\o
ller-Plesset perturbation theory (MP), the in-medium similarity
renormalization group (IM-SRG) method, coupled cluster (CC)
theory \cite{PhysRevB.67.045320,heidari:114708,PhysRevB.84.115302},
quasidegenerate perturbation theory (QDPT), and equation-of-motion
(EOM) methods.  We compare them against variational and diffusion
Monte Carlo (VMC and
DMC) \cite{PhysRevB.68.035304,PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780}
and full configuration interaction 
(FCI) theory \cite{olsen2013thesis,JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102}
results, where available,  from the existing literature.

@@

The article is organized as follows: Section \ref{sec:formalism}
introduces first (\ref{subsec:modelHamiltonian}) the Hamiltonian we
use to model circular quantum dots, and gives afterwards an overview
of the \textit{ab initio} many-body methods used in this paper:
Hartree--Fock, IM-SRG, CCSD, QDPT, EOM.  Our results are presented in
Section \ref{sec:results}.  We analyze and compare the differences
between the various methods and also between different quantum dot
systems.  We examine the usefulness of extrapolation techniques in
improving the precision of the results.  Section \ref{sec:conclusions}
concludes our work and gives perspectives for future work.

@@

We shall model the circular quantum dot system as a collection of $N$
nonrelativistic electrons of mass $m$ in two-dimensional space, trapped
by an external harmonic-oscillator potential of the form $m \omega^2r^2 / 2$, 
where $\omega$ is the angular frequency of the oscillator
trap and $r$ is the distance from the center of the trap.  The
electrons interact with each other through the standard Coulomb
interaction $e^2 / (4 \pi \epsilon R)$, where $e$ is the charge of
each electron, $\epsilon$ is the permittivity of the medium, and $R$
is the distance between the two interacting electrons.

Even though the model system contains five  parameters $(N, m, \omega,
e, \epsilon)$, many of them are redundant.  With an appropriate choice
of units, the number of parameters in the system can be reduced to
just two.  For rest of this paper, we shall use atomic
units
%\footnote{Here, ``atomic unit'' is used in an \emph{abstract}
%sense: we do not assume (nor care) whether $m$, $e$, $\epsilon$
%correspond to true values in vacuum, or some effective value in
%medium.} 
where $\hbar$, $m$, $e$ and $4 \pi \epsilon$ are all set to
one.  The unit Hartree $E_{\mathrm{h}}$ becomes the unit of energy and the Bohr
radius $a$ becomes the unit of length:
\begin{align*}
  E_{\mathrm{h}} &\equiv m \left(\frac{e^2}{4 \pi \epsilon \hbar}\right)^2 &
  a &\equiv \frac{4 \pi \epsilon \hbar^2}{m e^2}.
\end{align*}
This leaves us with $(N, \omega)$ as the only parameters necessary to
specify the quantum dot system, with $\omega$ in units of
$E_{\mathrm{h}} / \hbar$.

The behavior of each electron in isolation is described by the
nonrelativistic kinetic energy $t(\bm{r})$ and the potential energy
due to the trap $u(\bm{r})$, which combine to form the standard
harmonic oscillator Hamiltonian $h(\bm{p}, \bm{r})$:
\begin{align*}
  t(\bm{p}) &\equiv \frac{1}{2} \bm{p}^2, &
  u(\bm{r}) &\equiv \frac{1}{2} \omega^2 \bm{r}^2, &
  h(\bm{p}, \bm{r}) &\equiv t(\bm{p}) + t(\bm{r}).
\end{align*}
Here, $\bm{p}$ is the linear momentum of the electron and $\bm{r}$ is
its position relative to the center of the trap.  The pairwise
interaction between two electrons at positions $\bm{r}$ and $\bm{r}'$
is simply the Coulomb interaction:
\begin{align*}
  v(\bm{r}, \bm{r}') \equiv \frac{1}{|\bm{r} - \bm{r}'|}.
\end{align*}
Then, the full many-body problem is described by the Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{[1]} + \hat H^{[2]},
\end{align*}
where $\hat{H}^{[1]}$ and $\hat{H}^{[2]}$ are the one- and two-body parts of the Hamiltonian, respectively, and given 
\begin{align} \label{eq:onetwobodyhamiltonian}
\hat{H}^{[1]} &\equiv \sum_{j = 1}^N h(\hat{\bm p}_j, \hat{\bm r}_j), &
\hat{H}^{[2]} &\equiv \sum_{j_1 = 1}^N \sum_{j_2 = 1}^{j_1 - 1} v(\hat{\bm r}_{j_1}, \hat{\bm r}_{j_2}).
\end{align}
The operator $\hat{\bm r}_j$ is the position operator of the $j$-th
particle,\footnote{The ordering of the particle labels $j$ is
unimportant as they exist only for the purpose of bookkeeping.} and
$\hat{\bm p}_j \equiv -\mathrm{i} \hat{\bm{\nabla}}_{\bm{r}_j}$ is its
momentum operator.

@@

The noninteracting part of the many-body Hamiltonian $\hat{H}^{[1]}$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2.
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle,
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  In Cartesian
coordinates, the two-dimensional harmonic oscillator is trivially
reducible to the well-known one-dimensional problem.  However, to
exploit the circular symmetry, we prefer to use Fock-Darwin states,
which are written in polar coordinates $\bm{r} = (\rho, \varphi)$.
Such states have the computationally useful property of conserving
orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv
-\mathrm{i} \partial / (\partial \varphi)$.

The Fock-Darwin wave functions can be decomposed into radial and
angular components,\cite{lohne2010coupled}
\begin{align}
  F_{n, m_\ell}(\rho, \varphi) &\equiv \sqrt\omega R_{n,
  |m_\ell|}(\sqrt \omega \rho)
  A_{m_\ell}(\varphi) \label{eq:fockdarwin}, \\ R_{n, m}(\varrho)
  &\equiv \sqrt{\frac{2 \cdot n!}{(n + m)!}} \mathrm{e}^{-\varrho^2 /
  2} \varrho^m L_n^{(m)}(\varrho^2), \notag \\ A_{m_\ell}(\varphi)
  &\equiv \frac{1}{\sqrt{2 \pi}} \mathrm{e}^{\mathrm{i}
  m_\ell \varphi}, \notag
\end{align}
where $L_n^{(\alpha)}$ denotes the (generalized) Laguerre
polynomial \cite{NIST:DLMF} of degree $n$ and parameter $\alpha$,
\begin{align*}
  L_n^{(\alpha)}(u) = \frac{1}{n!} u^{-\alpha} \mathrm{e}^u \frac{\mathrm{d}^n}{\mathrm{d} u^n} (\mathrm{e}^{-u} u^{\alpha + n}).
\end{align*}
The states are labeled by two quantum numbers: the principal quantum
number $n \in \{0, 1, 2, \ldots\}$ and orbital angular momentum
projection $m_\ell \in \mathbb \{\ldots, 2, -1, 0, 1, 2, \ldots\}$.
For a wave function, $n$ indicates the degree of the Laguerre
polynomial, whereas $m_\ell$ is simply the eigenvalue of
$\hat{L}_{\mathrm{z}}$.

Since electrons are spin-$1/2$ fermions, they can occupy either of the
two possible spin states $\chi_{-1/2}$ or $\chi_{+1/2}$.  Thus, every
single-particle basis state $| (n, m_\ell, m_{\mathrm{s}}) \rangle$
contains both a spatial component \eqref{eq:fockdarwin} and a spin
component,
\begin{align} \label{eq:singleparticlestate}
  \langle (\rho, \varphi), m_{\mathrm{s}}' | (n, m_\ell,
  m_{\mathrm{s}}) \rangle \equiv F_{n,
  m_\ell}(\rho, \varphi) \delta_{m_{\mathrm{s}}^{}, m_{\mathrm{s}}'}.
\end{align}
Here we have introduced spin projection $m_{\mathrm{s}} \in \bigl\{-1/2, +1/2\bigr\}$ as the third quantum number and $\delta_{p, q}$ to denote the Kronecker delta..

The energy of the single-particle state $| (n, m_\ell, m_{\mathrm{s}}) \rangle$ is given by
\begin{align} \label{eq:energysingleparticlestate}
  \varepsilon_{n, m_\ell, m_{\mathrm{s}}} \equiv (2 n + |m_\ell| + 1) \omega.
\end{align}
The energies are degenerate with respect to the spin projection $m_{\mathrm{s}}$ as our Hamiltonian $\hat{h}$ does not distinguish between them.  Additionally, they are degenerate with respect to the nonnegative integer
\begin{align} \label{eq:shell_index}
  k \equiv 2 n + |m_\ell|,
\end{align}
which we conveniently call the \textit{shell index} as it labels each
shell starting from zero.  The shells are equidistant with an energy
spacing equal to the frequency $\omega$.

Returning to the one-body part of the many-body Hamiltonian,
$\hat{H}^{[1]}$, we can use the Slater determinant construction to
build an $N$-particle eigenstate $|p_1, \ldots, p_N\rangle$ of the
many-body Hamiltonian $\hat{H}^{[1]}$ from an arbitrary choice of $N$
single-particle eigenstates $|p_1\rangle, \ldots, |p_N\rangle$
from \eqref{eq:singleparticlestate}, where $p$ is an abbreviation for
the triplet $(n, m_\ell, m_{\mathrm{s}})$.  Note that a Slater
determinant $|p_1, \ldots, p_N\rangle$ is not merely a product of the
single-particle states, but an antisymmetrized sum over all possible
products,
\begin{align*}
  \langle \bm{r}_1, \ldots, \bm{r}_N | p_1, \ldots, p_N \rangle \equiv
  \frac{1}{\sqrt{N}} \left|
  \begin{matrix}
    \langle \bm{r}_1 | p_1 \rangle & \cdots & \langle \bm{r}_1 | p_N \rangle \\
    \vdots & \ddots & \vdots \\
    \langle \bm{r}_N | p_1 \rangle & \cdots & \langle \bm{r}_N | p_N \rangle \\
  \end{matrix}
  \right|.
\end{align*}
The Slater determinant $|p_1, \ldots, p_N\rangle$ is said
to \textit{occupy} the states $|p_1\rangle, \ldots, |p_N\rangle$.  By
construction, Slater determinants satisfy the Pauli exclusion
principle, ensuring that the exchange of any two particles is
equivalent to a reversal in the sign of the many-body wave function.

@@

In this work we will mainly work with the formalism of second quantization, see for example Ref.~\cite{shavitt2009many}.
To properly describe fermions, we introduce the annihilation and creation operators $\hat a_p$ and $\hat a_p^\dagger$. These operators must satisfy the anticommutation relations that encode the Pauli principle, namely
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0, &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p, q}.
\end{align*}

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from Eq.~\eqref{eq:onetwobodyhamiltonian} can be rewritten as
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{pq} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{}, &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{pqrs} \langle pq | \hat{H}^{[2]} | rs \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}.
\end{align}
where the quantities $\langle p | \hat{H}^{[1]} | q \rangle$
and $\langle pq | \hat{H}^{[2]} | rs \rangle$ are known
as \textit{matrix elements} and fully characterize the two operators
in the basis.  In this paper, we use the letters $p, q, r, s,
t, \ldots$ to label arbitrary single-particle \textit{states} in the
basis.  For the harmonic oscillator basis, labels such as $p$ can be
considered abbreviations for triplets of the form $(n, m_\ell,
m_{\mathrm{s}})$.  A summation such as $\sum_p$ is then understood as
a summation over every single-particle basis state $p$.  This is not
to be confused with summations such as $\sum_j$
from \eqref{eq:onetwobodyhamiltonian}, which contains a label $j$
denoting an arbitrary \textit{particle}.  We thus see another benefit
with second quantization: the Hamiltonian representation makes no
attempt to distinguish particles at all, as expected for particles
with quantum statistics.

The complete Hamiltonian is given by the sum
\begin{align}
  \hat{H} = \sum_{p,q} \langle p | \hat{H}^{[1]} | q \rangle \hat{a}_p^\dagger \hat{a}_q + \frac{1}{4} \sum_{pqrs} \langle pq | \hat{H}^{[2]} | rs \rangle \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r . \label{eq:vacuumhamiltonian}
\end{align}
Due to the antisymmetry of Slater determinants, we have the property
that exchanging either $p$ with $q$ or $r$ with $s$ reverses the sign
of $\langle pq | \hat{H}^{[2]} | rs \rangle$, leading to a 4-fold
symmetry that must be canceled out by the $1/4$ factor in front of the
summation in \eqref{eq:vacuumhamiltonian}.  For this reason, the
two-body matrix element $\langle pq | \hat{H}^{[2]} | rs \rangle$
is commonly referred to as the \emph{antisymmetrized} matrix element.
This is not to be confused with the (non-antisymmetric) interaction
integral,
\begin{align} \label{eq:interactionintegral}
  v_{p_1p_2p_3p_4} \equiv \int \langle p_1 | \bm{r} \rangle \langle p_2 | \bm{r}' \rangle v(\bm r, \bm r') \langle \bm{r} | p_3 \rangle \langle \bm{r}' | p_4 \rangle \D \bm{r} \D \bm{r}'.
\end{align}
This object only has a two-fold symmetry: $v_{p_1p_2p_3p_4} = v_{p_2p_1p_4p_3}$.  The antisymmetrized matrix element are related to the interaction integral via
\begin{align} \label{eq:antisymmetricmatrixelement}
  \langle pq | \hat{H}^{[2]} | rs \rangle = v_{pqrs} - v_{pqsr}
\end{align}
